2012 USA Presidential Election -- Part 1: Data Munging *Only* -- by Andrew Lavers
===================

Please see separate html report for the full project.


This project focuses on the 2012 USA presidential campaign finance data available from
http://fec.gov/disclosurep/PDownload.do. 

This project uses a *sample* from the [very large full countrywide dataset](ftp://ftp.fec.gov/FEC/Presidential_Map/2016/P00000001/P00000001-ALL.zip) to make analysis manageable. Using sampling rather than a single state data set enables some interesting state comparisons to be made.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library (reshape2)
library(plyr)
library(dplyr)
library (ggplot2)
library(lubridate)
```
### Munge Process Settings

These settings can be changed to make it easier to test the munge process. The data set  is large
so we recommend lower settings while exploring.
```{r munge_setting}
workdir <- "/Users/alavers/Documents/Udacity/Data Analysis with R/P3/"
# surce data file 
file_all    <- paste(workdir,"P00000001-ALL.csv" ,         sep="")
# number of rows from large file. Changing will invalidate randomness of sample
file_all_nrow <- -1
# raw sample file - to avoid resampling the very large file
file_sample <- paste(workdir, "P00000001-ALL_sampled.csv" ,sep="")
# munge result 
file_campaign <- paste(workdir, "campaign_5.csv" ,sep="")
# samppling settings
set.seed = 2341767
sample_size <- 0.05
setwd(workdir)
```

### Load  the data 

This data set has rows with a trailing comma on each row. This causes read.csv to make
assumptions because the header is shorter than the rows by 1. After much struggling
this Stack overflow link
http://stackoverflow.com/questions/13239639/duplicate-row-names-error-reading-table-row-names-null-shifts-columns
gave me the hint. 

The approach to read skips the header by using skip=1, and provides the column names - with one "extra" column -  
using the col.names attribute.

```{r warning=FALSE, load_data}
# the file has trailing commas on data rows making the col names short by 1 column
# So, read column names separately, add one more dummy column column "dummy"
ln <- readLines(file(file_all), n=1) 
ln <- paste (ln,'dummy', sep=',')               
col_names_raw <- unlist(strsplit(ln, ","))

#read csv skipping the first row and supplying the col names
campaign_raw <- read.csv(file_all, col.names=col_names_raw, header=FALSE, row.names=NULL,
                         skip=1, nrow=file_all_nrow)
campaign_sample <- sample_frac(campaign_raw, sample_size)
write.csv(campaign_sample, file_sample, row.names=FALSE)
```
### Check the counts and some sanity checks on the sample vs. full file
```{r warning=FALSE, echo=FALSE, counts}
print(paste("Rows read from all file:",nrow(campaign_raw)))
print(paste("Rows written to sample file:",nrow(campaign_sample)))

# sanity check totals for raw file ...
rep_count = nrow(subset(campaign_raw, !(cand_nm %in% c('Obama, Barack', 'Stein, Jill'))))
rep_sum = sum(subset(campaign_raw, !(cand_nm %in% c('Obama, Barack', 'Stein, Jill')), select = contb_receipt_amt))
tot_count <- nrow(campaign_raw)
tot_sum <- sum(campaign_raw$contb_receipt_amt)
print(paste("Republican FULL FILE count%:", 100 * rep_count / tot_count, " sum%: ", 100 * rep_sum / tot_sum))
summary(campaign_raw[,c('cand_nm','contb_receipt_amt','contbr_st')])

# ... and for sample
rep_count = nrow(subset(campaign_sample, !(cand_nm %in% c('Obama, Barack', 'Stein, Jill'))))
rep_sum = sum(subset(campaign_sample, !(cand_nm %in% c('Obama, Barack', 'Stein, Jill')), select = contb_receipt_amt))
tot_count <- nrow(campaign_sample)
tot_sum <- sum(campaign_sample$contb_receipt_amt)
print(paste("Republican SAMPLE count%:", 100 * rep_count / tot_count, ", sum%:", 100 * rep_sum / tot_sum))
summary(campaign_sample[,c('cand_nm','contb_receipt_amt','contbr_st')])
```
Based on the above simple stats the sample appears to be a good representation of the full data set.

### Rerun point

When rerunning we can just run from the sample read to make it quicker
``` {r rerun_point}
campaign_sample <- read.csv(file_sample, row.names=NULL)
```
### Drop unused columns 
```{r slice_colums}
# Slice each dataset by removing unwanted columns
campaign_tmp1 <- subset(campaign_sample, TRUE,
                        select = - c(memo_text, receipt_desc, file_num, memo_cd, tran_id, dummy,
                                     cmte_id, cand_id, form_tp))
```

### Refunds 

Remove the refunds, which make up about 3-4%. The matching contributions
should really be backed out as well but hopefully they are spread across the data set
and wont influence the result. A more rigorous analysis should clear this. With low sampling rate there is almost no chance that both the contribution and refund will be in the same sample. 
``` {r refunds}
# total amount of refunds vs total is small so  just drop refunds
total_refunds  <- sum(subset(campaign_tmp1,contb_receipt_amt <=0,contb_receipt_amt))
total_contribs <- sum(subset(campaign_tmp1,contb_receipt_amt >0,contb_receipt_amt))
percent_refunds <- total_refunds / total_contribs 
print(paste("Refunds:", total_refunds, "Contributions:", total_contribs, "Proportion refunds:", percent_refunds))

campaign_tmp2 <- subset(campaign_tmp1,contb_receipt_amt >0)
```

### Limit to $2,500, no dates before January 1, 2011, drop Green Party

Investigate excluding large amounts large amounts. They mostly appear to be transfers form other campaign funds and some corrections and company or fund contributions to the $5000 limit.
Individual limit per candidate is $2500, so lets stick to this subset. Note we use right=T to make the break numbers like 25, 50 land in the previous 
bucket. There are very few contributions to the green party, so exclude them to
make the parties comparable,.

```{r eliminate_large_amounts}
# year month of the date
campaign_tmp2$contb_receipt_ym<-strftime(dmy(campaign_tmp2$contb_receipt_dt),"%Y-%m")

# limit to $2500 and under and scrub some old dates and filter out the Gree Party (Jill Stein)
campaign_tmp3 <- subset(campaign_tmp2,contb_receipt_amt <=2500 &
                        contb_receipt_ym >= '2011-01-01' &
                        cand_nm != "Stein, Jill")
summary(campaign_tmp3$contb_receipt_amt)
```

Clearly there are many small contributions

### Filter States 

There are some odd state codes for the territories and the armed forces and the District of Columbia. 
We will only analyze the 50 states because these are electoral college states that actively participate in the 
election of the president.

```{r exclude_some_state_codes}
states <- c('AL','AK','AZ','AR','CA','CO','CT','DE','FL','GA','HI','ID','IL','IN','IA','KS','KY','LA','ME','MD','MA','MI','MN','MS','MO','MT','NE','NV','NH','NJ','NM','NY','NC','ND','OH','OK','OR','PA','RI','SC','SD','TN','TX','UT','VT','VA','WA','WV','WI','WY')

campaign_tmp4 <- subset(campaign_tmp3, contbr_st %in% states)
campaign_tmp4$contbr_st <- factor(campaign_tmp4$contbr_st)
```

### Augment with Swing State and Politcal Party

Add 
- *Swing state*  as sourced from http://elections.nytimes.com/2012/swing-state-tracker 

- *Political party* information from various Wikipedia entries.

- *5-digit zip* from first five of 9-digit zip

- *Year-month* from receipt date to simplify month-by-month plotting

```{r state_columns}
# augment with swing states
swing_states = c('IA', 'NH', 'WI', 'FL', 'VA', 'NV', 'OH', 'CO', 'NC')
campaign_tmp4$swing_st <- campaign_tmp4$contbr_st %in% swing_states

# augment with party information (retrieved from various Wikipedia entries)
cand_nm <-
          c("Obama, Barack", "Romney, Mitt", "Bachmann, Michele", "Santorum, Rick", "Paul, Ron", 
            "Gingrich, Newt", "Perry, Rick", "Cain, Herman","Pawlenty, Timothy", "Huntsman, Jon", 
            "Johnson, Gary Earl", "Roemer, Charles E. 'Buddy' III", "Stein, Jill", "McCotter, Thaddeus G")
party <- 
          c("Democratic", "Republican", "Republican", "Republican", "Republican",
            "Republican", "Republican", "Republican", "Republican", "Republican",
            "Republican", "Republican", "Green", "Republican")
cand_party <- data.frame(cand_nm, party)
campaign_tmp5 <- merge(campaign_tmp4, cand_party)

# zip 5
campaign_tmp5$contbr_zip5 <- substr(campaign_tmp5$contbr_zip,start=1, stop=5)
``` 

### Write the final dataframe to csv file for analysis
```{r final_data_frame}
# final base data frame
campaign <- campaign_tmp5
write.csv(campaign, file_campaign, row.names=FALSE)
```