---
output:
  html_document:
    fig_width: 7
    fig_height: 5
---
2012 US Presidential Election Campaign Contributions – Part 2: by Andrew Lavers
========================================================

The 2012 presidential election was on by Barack Obama of the Democratic Party running aginst Mitt Romney of the Republican party. The genrela election followed the earlier primary election phase which was contested in the Republican party. No Democrats contested Obama in a primary for the Democratic nomination.

Campign finance regulations require the recording and publishing of
campaign contributions. This report describes an analysis of that
data.

This project uses a *sample* from the [very large full countrywide FTP dataset](ftp://ftp.fec.gov/FEC/Presidential_Map/2016/P00000001/P00000001-ALL.zip) to make analysis manageable. Using sampling rather than a single state data set enables some interesting state comparisons to be made. All data below is reported for the SAMPLE, and has NOT been adjusted for the full dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(lubridate)
library(scales)
library(gridExtra)
options(scipen=20)
workdir <- "/Users/alavers/Documents/Udacity/Data Analysis with R/P3/"
setwd(workdir)
file_campaign <- paste(workdir, "campaign_5.csv", sep="")
file_session  <- paste(workdir, "P3Session.Rda",  sep="")
sample_frac <- 0.05
sample_perc <- 100*sample_frac


# -------- REUSABLE FUNCTIONS AND VARIABLES

# some colors for later
count_color <- 'goldenrod'
count_alpha <- 0.5
amt_color <- 'darkgreen'
amt_alpha <- 0.7
party_colors <- c('blue', 'red')
party_alpha <- 0.5

# #modified paired palette to avoid light yellow
# simply switch the last two pairs (colorbrewer.or)
paired_palette <- c(
                     '#1f78b4', '#a6cee3',  
                     '#33a02c', '#b2df8a',
                     '#e31a1c', '#fb9a99', 
                     '#ff7f00', '#fdbf6f',
                     '#6a3d9a', '#cab2d6',
                     '#b15928', '#ffff99')


# define a commmon date scale formatting
scale_x_date1 <- scale_x_date (breaks = '3 months', minor_breaks = '1 month',
                     labels = date_format("%b '%y"),
                     limits = c(as.Date('2011-01-01'), 
                                as.Date('2012-11-30') ) )
scale_x_date_prim <- scale_x_date (breaks = '6 months', minor_breaks = '3 month',
                     labels = date_format("%b '%y"),
                     limits = c(as.Date('2011-01-01'), 
                                as.Date('2012-08-30') ) )

# define labelsc formatter for text yyyy-mm as mmm 'yy
labels_mmm_yy <- function  (yyyy_mm) {
  formatted <- strftime(ymd(paste(yyyy_mm, "15", sep="-")), "%b '%y")
  return (formatted)
}
# define breaks for text yyyy-mm 
breaks_yyyy_mm <- function (yyyy_mm, every=3, ending_at=11) {
  # break on 3 months, relative to 11 November (the election month)  
  mm <- as.integer(substr(yyyy_mm, 6, 7))
  return (ifelse((mm %% every) == (ending_at %% every),
                  yyyy_mm, 
                  "") )
}

# for year month, define a common scale
scale_x_yyyy_mm = scale_x_discrete(breaks = breaks_yyyy_mm, 
                                   labels = labels_mmm_yy)

# define common theme tweaks
theme1 <- theme(title = element_text(vjust = 1.5),
                axis.title.x = element_text(vjust = -0.5),
                axis.title.y = element_text(hjust = 0.5) )
theme_x_state <- theme(axis.text.x = element_text(angle = 90, 
                                                  vjust = 0.4,
                                                  size = 8),
                       panel.grid.major = element_blank())

# define function for primary candidates
# this is not very robust becaseu it assuems all other are priamry
isPrimaryCandidate <- function (cand_nm) {
  return (!(cand_nm %in% c("Obama, Barack","Romney, Mitt")))
}

# define a function to exclude dates after the primary
# not very robust becasue it assumes all other dates
# are prinary
isPrimaryYm <- function (yyyy_mm) {
# months in the data that are NOT the primary elections
  return (!(yyyy_mm %in% c('2012-09', '2012-10', '2012-11', '2012-12')))
} 

```
### Load the data 
IMPORTANT: This is a `r sample_perc`%  sample from the complete data set. 
This should be sufficient to represent the full data set but actual totals 
of contributions will not represent the totals. Any totals reported here have 
NOT been adjusted for sampling.

The data to be loaded is in file *campaign_5.csv*. This is a munged data set based on
the presidential campaign ALL states data set. The munging is documented in the separate
[file:AndrewLaversCampainMunge.html] produced from [file:AndrewLaversCampainMunge.Rmd]

```{r echo=FALSE, warning=FALSE, Load_the_Data}
# Load the Data 
campaign <- read.csv(file_campaign, row.names=NULL)

# make a date type from the month text using yyyy-mm-15 - midpoint 
campaign$contb_receipt_ym_date <- ymd(paste(campaign$contb_receipt_ym, 
                                            "15", 
                                            sep = '-'))
campaign$contb_receipt_dtd <- dmy(campaign$contb_receipt_dt)
# cutoff in November the election month. There are some contributions
# in December but they are small and distort the charts

campaign <- subset(campaign, contb_receipt_dtd < ymd('2012-12-01'))

# factors
campaign$party <- factor(campaign$party, 
                         levels = c("Democratic","Republican"))
campaign$swing_st <- factor(campaign$swing_st, 
                            levels = c(TRUE, FALSE), 
                            labels = c("Swing State", "Non-swing State"))
amt_breaks <- c(0, 25, 50, 100, 250, 1000, 2500)
campaign$contb_receipt_amt_category <- cut(campaign$contb_receipt_amt, 
                    breaks=amt_breaks,
                    labels=c('(0,25]',    '(25,50]',    '(50,100]', 
                             '(100,250]', '(250,1000]', '(1000,2500]') )

print (paste ("Analyzing ", nrow(campaign), "rows from", file_campaign))
```

----------------------------------

# Univariate Plots Section

### Contribution Amounts

Investigate contribution amounts and establish a category that will
make analysis of contributions more consistent.

```{r echo=FALSE, warning=FALSE, Univariate_Plots1a}
# histogram of amount
ggplot(campaign, 
       aes(x = contb_receipt_amt)) +
  stat_bin(breaks = seq(0,2500,25), 
           right = T, 
           fill = amt_color, 
           alpha = amt_alpha) +
  scale_x_continuous(breaks = seq(0,2500,200)) +
  scale_y_sqrt(breaks = c(100,1000,5000,10000,50000)) +
  labs(title="Histogram of Contribution Amounts",
       x = "Contribution Amount in $",
       y = "Count") +
  theme1
```

  At first thought, contribution amount is a continuous variable, but in 
  reality there are very distinct buckets that show in the chart as spikes 
  at 25, 50 100, 250, 500, 1000, 1500, 2000, 2500.
  
  
  Clearly there are many small contributions, which is confirmed by the 
  following basic statistics of the contribution amount.

```{r echo=FALSE, warning=FALSE, Univariate_Plots1b}
summary (campaign$contb_receipt_amt)
```

  Based on the above chart and some experimentation, 
  the contribution was encoded in the distinct 
  buckets as a new category variable named *contb_receipt_amt_category*.
  
The levels of this new category valriable can be clearly seen in the followign chart:

```{r echo=FALSE, warning=FALSE, encode_buckets}
ggplot(campaign, 
       aes(x=contb_receipt_amt_category)) +
  scale_x_discrete() +
  geom_histogram(fill = count_color, 
                 alpha = count_alpha ) +
  labs(title="Contribution Counts by Contribution Amount Category",
       x = "Contribution Amount Category in $",
       y = "Count") +
  theme1

```
  
  After some experimentation the above buckets produced the expected 
  falling counts by increase in contribution amount. As is expected, 
  the number of contributors declines as the contribution amount 
  increases which reflects the overall country wealth demographics. 
  
----

### Contributions by Candidate

A basic look at the difference by candidate.

```{r echo=FALSE, warning=FALSE, Univariate_Plots2}
# histogram by candidate
ggplot(campaign, 
       aes(x = reorder(cand_nm, cand_nm, length)))  +
  geom_histogram(fill = count_color, 
                 alpha = count_alpha) +
  scale_y_log10() +
  labs(x = "Candidate", 
       y = "Contribution Count",
       title="Contribution Counts by Candidate") +
  coord_flip() +
  theme1
```

  In the above chart, the two final presidential candidates for the 
  general election are orders of magnitude greater than the primary 
  candidates, so further analysis of the primary candidates is not 
  likely to be that interesting.

----

### Contributions over time

Every contribution in the data set has the date of the contribution so
we can do some analysis of the contributions over time. 

```{r echo=FALSE, warning=FALSE, Univariate_Plots3}
# histogram by year-month
ggplot(campaign, 
       aes(x = contb_receipt_ym) ) +
  geom_bar(
           fill = count_color, 
           alpha = count_alpha) +
  scale_y_sqrt(breaks = c(100,1000,5000,10000,20000,40000,80000)) +
  scale_x_yyyy_mm +
  labs(title = "Contribution Counts by Time",
       x = "Year - Month",
       y = "Count") +
  theme1
```

  Contributions accelerate approaching the early November election date,
  as can be seen in the above chart. It will be interesting to see the 
  different pace of Republican versus Democratic contributions. 

----

### Contributions by State

States are an important factor in the presidential election because
of the way the Electoral College of state representatives actually
elects the president.

```{r echo=FALSE, warning=FALSE, Univariate_Plots4}
# histogram by state using a very hacky way of 
# ordering by count - there's gotta  be a better way!!!
ggplot(campaign, 
       aes(x=reorder(contbr_st, contbr_st, length ))) +
  geom_histogram(fill = count_color, 
                 alpha = count_alpha) +
  scale_y_continuous(expand = c(0,0)) +
  labs(title = "Contribution Counts by State",
       x = "State",
       y = "Count") +
  theme_x_state +
  theme(panel.grid.major.x = element_blank()) +
  theme1
```
  
  The above counts by state show a few large states as frequent 
  contributors. 

Swing states are important because the president is elected through 
the Electoral College. States that are evenly split between the parties 
can "swing" the election because they vote all their Electoral votes 
based on the results with in the estates. See 
[2012 swing states](http://elections.nytimes.com/2012/swing-state-tracker)


``` {r echo=FALSE, warning=FALSE, Univariate_Plots6}
# counts by swing state
ggplot(campaign, 
       aes(x=swing_st)) +
  geom_histogram(fill = count_color, 
                 alpha = count_alpha,
                 width = 0.3) +
  labs(title = "Contribution Counts by Swing vs Non-Swing State",
       x = "",
       y = "Count") +
  theme1 +
  theme(legend.position = 'none',
        panel.grid.major.x = element_blank())

```

  While the above chart shows that there are substantial differences 
  in contributions when comparing Swing States to Non-swing states, 
  the overall state populations are most likely masking the effect. 
  Very populous states such as NY, CA and TX are not swing states.

----

### Contributions by Party

A basic investigation of party differences.

``` {r echo=FALSE, warning=FALSE, Univariate_Plots5}
# counts by party
ggplot(campaign, 
       aes(x=party, 
           fill=party)) +
  geom_histogram(width=0.3, 
                 alpha=party_alpha) +
  scale_fill_manual(values=party_colors) +
  labs(title = "Contribution Counts by Political Party",
       x = "Party", 
       y = "Count") +
  theme1 +
  theme(legend.position = 'none',
        panel.grid.major.x = element_blank())
```

  The counts by party above are substantially different. We should later
  investigate the relative size and number of contribution by party. 

----

# Univariate Analysis

### What is the structure of your dataset?

The data set to analyze has individual contributions to 2012 presidential 
candidates. Each contribution has:

- *Contributor* - Name and address

- *Contributor* -  Occupation (Not useful because it isn't normalized. 
Many different equivalent. There may be a few interesting high-frequency 
common items like LAWYER, PHYSICIAN)

- *Contribution* -  Date, Year month, amount 

- *Commitee and Candidate* -  Committee ID, Candidate ID and Candidate Name 

- *Election type* - P2012 for primaries and G2012 for general election

- *Form type and transaction id* - Not used in this analysis

- *Party Affiliation* - Republican or Democratic

```{r echo=FALSE, structure}
```

### What is/are the main feature(s) of interest in your dataset?

- *Individual contribution* - Each individual contribution is represented 
as a row so the contribution counts 

- *Candidate (cand_nm)* - Obviously the presidential election is about 
the person that will be president

- *State (contbr_st)* - In presidential elections, states vote with Electoral 
college ballots, so votes within a state matter. See for example 
http://en.Wikipedia.org/wiki/Electoral_College_%28United_States%29. 
The state identifiers in this data set includes identifiers for non 
voting territorial possessions (e.g Guam, US Virgin Islands)

- *Contribution Amount (contb_receipt_amt)* - The is the dollar amount 
of the contribution is the most interesting item to analyze. This part 
of the data set was limited to include contribution under the 2012 
contribution limit of $2500. There varying reports of whether contributions 
under $200 must be reported. About half the contributions in this data set 
are $50 and under. We can analyze the differences and totals of contributions 
to reach broad conclusions, but these will not represent the full population 
of contributions.

### What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?

- *Date (contb_receipt_dt, contb_receipt_ym)* - Looking at contributions 
over time may prove interesting. The Republicans had a large primary 
field while the Democrats presented a single candidate, the incumbent 
president. Time may show us when the primary candidates dropped out.

- *Occupation (contbr_occupation)* - While this may be interesting, 
the values are not normalized and these cant be effectively compared across
all occupations. There are a few discrete occupations that may be 
interesting such as LAWER, PHYSICIAN, TEACHER.

- *Employer (conbr_employer)* - This will be very interesting to find 
employers with many contributors. However this may be better left for 
other type of analysis -- visualization with plots may not effectively 
tease out isolated hot spots. One category appears here "RETIRED", 
"HOMEMAKER", "UNEMPLOYED", "NOT EMPLOYED" that may be useful for broad 
categories but there are many "INFORMATION REQUESTED" which indicates 
missing data.

### Did you create any new variables from existing variables in the dataset?

- *Party (party)* - Party  helps pool the early contributions that went 
to multiple Republican primary candidates, into two main buckets 
Republican vs. Democrat. The party value was determined from Wikipedia 
articles and merged onto the main data set. 

- *Swing State (swing_st)* - In USA presidential elections, the 
president is actually elected 
by an Electoral college of state representatives who cast a preallocated 
number of electoral votes on behalf of their state. In many 
states, the full block of electoral votes must go to the winning 
candidate in that state. If these states have almost equal Democratic and 
Republican support, they can be the one state that "swings" the election. 
In 2012 there were 9 swing states as tracked by the 
[New York Times](http://elections.nytimes.com/2012/swing-state-tracker)

- *Categorized Contribution Amount (contb_receipt_amt_category)* - The 
charts show distinct levels 
of contribution at these breaks:  $25, $50, $100, $250, $1000, and $2500. 
A categorized variable was added to facilitate analysis.
  
- *Receipt year month (contb_receipt_ym)* - The Year month of the 
receipt date to simplify trend plotting.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

For more detail, see the data munging documented  file:AndrewLaversCampaignMunge.html

The main operations performed where:

- Remove negative values that represent returns. Strictly we 
should also find the original contribution and remove that as well.  
Finding the original is made more difficult by sampling. But since 
returns make up about 3% of the data set, this was left for a future 
exercise. Negative return values distort the means and medians

- Limit to contributions of $2500. This eliminated a few very 
large party committee transfers, 
many smaller corporate contributions and leaves a more consistent 
data set of individual contributions only. 

- Categorized the contribution amount into buckets

- Eliminate "states" that are territorial possessions etc., 
that do not form the Electoral College.

- Limit to dates after 1/1/2011

- Eliminate the Green Party because there are very few contributions 

----

# Bivariate Plots Section

### Parties vs. States

Here we look at some of the relationships between parties and states.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots5}
ggplot(campaign, 
       aes(x=party, 
           y = contb_receipt_amt, 
           fill=party)) +
  geom_boxplot(alpha=party_alpha) +
  facet_wrap(~swing_st) +
  scale_y_sqrt(breaks=amt_breaks, 
               expand = c(0,0)) +
  scale_fill_manual(values=party_colors) +
  labs(title = "Distribution of Contribution Amounts by Political Party\nSwing and Non-swing States",
       x = "Party", 
       y = "Contribution Amount in $") +
  theme1 +
  theme(legend.position = 'none')
```

As can be seen in the above chart, there is not much difference 
in individual contribution amounts in Swing or Non-swing states. 
The median Democratic Swing State contribution of 
`r median(unlist(subset(campaign, party=="Democratic" & swing_st == "Swing State", select=contb_receipt_amt)))` 
differs slightly from the 
`r median(unlist(subset(campaign, party=="Democratic" & swing_st == "Non-swing State", select=contb_receipt_amt)))`
 we find in Non-swing States.
 The Republican median of 
 `r median(unlist(subset(campaign, party=="Republican" & swing_st == "Swing State", select=contb_receipt_amt)))`
 is unchanged.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots6}
# bar of sum by state
campign_gr_st <- campaign %>% 
                 group_by(contbr_st, swing_st) %>%
                 summarize(sum_contb_receipt_amt = sum(contb_receipt_amt))
ggplot(campign_gr_st, 
       aes(x = reorder(contbr_st, sum_contb_receipt_amt, sum),        
           fill = swing_st)) +
  geom_bar(aes (y = sum_contb_receipt_amt/1E6),
           stat='identity',  
           alpha=amt_alpha) +
  scale_y_sqrt(breaks = pretty_breaks(), expand = c(0,0) ) +
  scale_fill_manual(values = c(amt_color, 'lightgreen')) +
  labs(title="Total Contribution Amount by State",
       x = 'State', 
       y = "Contribution Amount in $ millions") +
  theme_x_state +
  theme(legend.position='bottom', 
        legend.title = element_blank()) 
```

A SQRT scale is needed in the above chart to show the distribution. 
Clearly the populous states such as NY, CA, TX, FL, and  IL 
dominate in election contributions.

A cursory review of 
[US State populations](http://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population) 
suggests that the above chart corresponds approximately 
to state populations. Virginia, a swing state, may be an exception 
being 7th in contributions but only 12th in population.

Thus swing states don't seem to make an obvious difference.   

----

### Contributions by Candidate

The general election candidates,  Mitt Romney and Barack Obama will
obviously raise more because they are in the race to the end. 
These next charts investigate the differences between 
candidates, first by contribution size and then by total amount.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots7}
# boxplot
ggplot(campaign, 
       aes(x = reorder(cand_nm, contb_receipt_amt, sum),  
           y = contb_receipt_amt)) +
  geom_boxplot() +
  coord_flip() +
  labs (x="Candidate", 
        y="Per Contribution Amount in $", 
        title="Contributions by Candidate") + 
  theme1
```

The above chart shows significantly higher median contribution 
amounts for the primary candidates.  Rick Perry's median of 
 `r median(unlist(subset(campaign, cand_nm == "Perry, Rick", select=contb_receipt_amt)))`
is similar to that of Timothy Pawlenty and much higher than the 
rest of the field. Pawlenty has far fewer overall contributions 
so this similarity here may be deceiving. The difference in 
distribution between Mitt Romney and Barack Obama can clearly be seen. 

```{r echo=FALSE, warning=FALSE, Bivariate_Plots8}
# bars for total amount
campaign_cand <- campaign %>% 
                 group_by (cand_nm) %>%
                 summarize (sum_contb_receipt_amt = sum(contb_receipt_amt))
ggplot(campaign_cand, 
       aes(x = reorder(cand_nm, sum_contb_receipt_amt),
           y = sum_contb_receipt_amt/1E6)) +
  geom_bar(stat = "identity") +
  scale_y_sqrt() +
  coord_flip() +
  labs(title = "Total Contributions by Candidate",
       x = "Candidate", 
       y = "Total Contribution in $ millions" ) +
theme1
```

This result above is interesting to me because I have always 
wondered how much money was "wasted" during
primary elections. In the chart you can see that both 
Mitt Romney and Barack Obama received orders of magnitude  
more that the other primary-only candidate. 

----

### Contributions over time

We will now investigate contributions over time and how they relate to 
other factors.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots9A}
#bars for total contributions time
ggplot(campaign, 
       aes(x = contb_receipt_ym, 
           y = contb_receipt_amt/1E6)) +
  geom_bar(stat="identity")  +
  facet_wrap(~party, ncol = 1) +
  scale_x_yyyy_mm + 
  labs(title = "Total Contributions by Month",
       x = "Month",
       y = "Total Contributions in $millions")  +
  theme1

```

In the above  chart contributions accelerate as the election 
approaches, and stop immediately after. There is distinctly earlier 
contributions to the Republican Party between 8/20111 and 4/2012. 
The Democratic Party ends very strong with noticeably higher contributions.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots9B}
# get cumulative sums
campaign_party_dt <-  campaign %>%
                      filter(contb_receipt_dtd > ymd("2012-05-31")) %>%
                      group_by (party, contb_receipt_dtd) %>%
                      summarize(sum_amt = sum(contb_receipt_amt),
                                count = n()) %>% 
                      mutate(cumsum_amt   = cumsum(sum_amt),
                             cumsum_count = cumsum(count)) 

#plot lines for total amounts
ggplot(campaign_party_dt, 
       aes(x = as.Date(contb_receipt_dtd), 
           y = cumsum_amt/1e6, 
           color=party)) +
  geom_line()  +
  scale_color_manual(values = party_colors) +
  scale_y_continuous(breaks = pretty_breaks()) +
  scale_x_date(breaks = pretty_breaks() ) +
  labs(title = "Cumulative Contributions\nPeriod after Primary Election",
       x = "2012", 
       y = "Cumulative Amount in $Millions") +
  theme1

```

In the months after  the primaries, total contributions are 
similar until September when the Democratic Party jumps ahead by 
$2million. This lead is then maintained. The Democratic Party 
convention was held September 3-6, which could be a trigger for 
this additional contribution. Note that because of sampling at 
`r sample_perc`% this gain would be about $40 million. 

```{r echo=FALSE, warning=FALSE, Bivariate_Plots9C}
#plot lines for counts
ggplot(campaign_party_dt, 
       aes(x = as.Date(contb_receipt_dtd), 
       y = cumsum_count, 
       color = party)) +
  geom_line()  +
  scale_color_manual(values=party_colors) +
  scale_y_continuous(breaks = pretty_breaks()) +
  scale_x_date(breaks = pretty_breaks() ) +
  labs(title = "Cumulative Contribution Counts by Month\nPeriod after Primary Election",
       x = "2012", 
       y = "Cumulative Contribution Count") + theme1 
```

The Democratic Party leads dramatically in number of contributions 
after the end of the Republican Primaries- more than 100,000 
in this `r sample_perc` sample

```{r echo=FALSE, warning=FALSE, Bivariate_Plots9D}
#plot lines for ratio of counts
campaign_party_dt_counts <- dcast(campaign_party_dt, 
                                  contb_receipt_dtd ~ party, 
                                  value.var = 'cumsum_count')
campaign_party_dt_counts$ratio <- campaign_party_dt_counts$Democratic / 
                                  campaign_party_dt_counts$Republic
campaign_party_dt_counts <- subset(campaign_party_dt_counts, 
                                   !is.na(campaign_party_dt_counts$ratio))
ggplot(campaign_party_dt_counts, 
       aes(x = as.Date(contb_receipt_dtd), 
           y = ratio)) +
  geom_smooth(method = 'loess')  +
  scale_x_date() +
  labs(title = "Ratio of Cumulative Contribution Counts by Month\nPeriod after Primary Election",
       x = "2012", 
       y = "Ratio Democratic to Republican") +
  theme1
```

In the above chart the Democratic acceleration is clear with the 
ratio of cumulative contribution rising rapidly to more than 4:1. This
is perhaps one of the clearest conclusions of the analysis.

----

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

A clear relationship is between contribution amounts and 
contribution size which varies by party. The Democratic Party 
is the clear leader by contribution counts and the Republican party 
leads in contribution size. As may be expected contributions accelerate 
closer to the election. An interesting note is the so-called 
"convention bump". The press and attention around the national 
convention attracts more interest and contributions. This bump 
in September 2012 can be clearly seen for the Democratic Party.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

There seem to be some similarities between the primary candidates who 
ultimately dropped out in the primaries. These can be seen in the box 
plot by candidate above. In particular the contribution size median is 
much higher suggesting that in order to progress in a primary election, 
the candidate must be supported by large contributions. In the next 
section we will explore this further and one of the final plots will 
show this relationship.

### What was the strongest relationship you found?

The strongest relationship is between party and contribution count and 
size. Both parties raised similar total amounts.  
The Democratic Party contribution count was about four times 
that of the Republican Party. 

----

# Multivariate Plots Section

### Cumulative Contribution by Party 

The next charts focus on cumulative sums to investigate the rate of contributions over time. The slope of a cumulative chart indicates the rate of change. 

Firstly, a chart to investigate which contribution size dominated in each party and how that varied over time

```{r echo=FALSE, Multivariate_Plot1}
# Compare amounts conntributed over time by party (excludign the green party)
campaign_gr_party_ym <- campaign %>% 
               group_by(party, 
                        contb_receipt_amt_category, 
                        contb_receipt_ym_date) %>%
               summarise(
                         sum_amt = sum(contb_receipt_amt), 
                         mean_amt = mean(contb_receipt_amt),
                         median_amt = median(contb_receipt_amt),
                         max_amt = max(contb_receipt_amt),
                         count=n()) %>%
               mutate(cumsum_amt = cumsum(sum_amt))

ggplot(campaign_gr_party_ym, 
      aes(x = as.Date(contb_receipt_ym_date), 
          y = cumsum_amt/1E6, 
          color=contb_receipt_amt_category,
          fill=contb_receipt_amt_category,
          group=contb_receipt_amt_category)) +
  geom_area(position = 'stack', alpha = 0.2) +
  scale_x_date1 +
  scale_color_discrete (name = "Contribution\nCategory $") +
  scale_fill_discrete  (name = "Contribution\nCategory $") +
  facet_wrap(~party, ncol=1)  +
  labs (title="Categorized Contribution Amounts by Category/Month",
        x = "", 
        y = "Total Contribution in $million's") +
  theme1 + 
  theme(legend.position = 'bottom')
```
 
  The above area plot reveals 
  significant differences between Republican and Democratic donation 
  patterns over time. The Republican Party reached $10 million in May 2012 
  with about $6 million coming from the top category. The Democratic Party reached
  the $10 million mark two months later with almost even contribution totals in 
  the top three categories.  The Republican chart is dominated by the top 
  contribution category.
  
  Next, we are interested in the rate of contributions, how fast they 
  grew, when they started and when they ended up. We will omit Mitt 
  Romney and Barack Obama because their totals are much greater than 
  the other candidates.
 
```{r echo=FALSE, warning=FALSE, Multivariate_Plots2}
# Create cumulative sums and other stats by candidate by month
primary_gr_nm_ym <- campaign %>% 
               filter(isPrimaryCandidate(cand_nm) &
                     isPrimaryYm(contb_receipt_ym)) %>%
               group_by(cand_nm, 
                        contb_receipt_ym) %>%
               summarise(sum_amt = sum(contb_receipt_amt), 
                         mean_amt = mean(contb_receipt_amt),
                         median_amt = median(contb_receipt_amt),
                         max_amt = max(contb_receipt_amt),
                         count=n()) %>%
               mutate(cumsum_amt = cumsum(sum_amt))

# plot cumulative amounts for primary
ggplot(primary_gr_nm_ym, ) +
  geom_line(aes(x = contb_receipt_ym, 
                y = cumsum_amt/1E6, 
                group=cand_nm)) +
  scale_x_yyyy_mm +
  facet_wrap(~cand_nm) +
  scale_color_brewer(palette="Paired") + 
  labs(title = 'Cumulative Contribution Amounts for Primary Election', 
       x = 'Month', 
       y = 'Total Contribution in $millions') +
  theme1 +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

  The faceted plot above  is interesting because it shows very 
  different shapes for non-starters and those that remained competitive. 
  We will  improve this for the for the final plots

--- 

### Contribution Size by State 

Does residing in a swing state influence the contribution size? 
The next charts explores relationships with the state of residence.

```{r echo=FALSE, warning=FALSE, Multivariate_Plots3}
ggplot(campaign, 
       aes(x=contbr_st, 
       y=contb_receipt_amt_category, 
       color=swing_st)) +
  geom_jitter(alpha=0.05) +
  labs(x = "State", 
       y = "Contribution Category",
       title = "Contribution Size by State") +
  scale_color_manual(values=c(amt_color, 'lightgreen')) +
  scale_y_discrete(expand=c(0,0)) +
  # low alpha causes faded legends, so override
  guides(size = FALSE, 
         color = guide_legend(override.aes = list(alpha = 1, size=6))) +
  theme_x_state +
  theme1 +
  theme (legend.position = 'bottom',
         legend.title = element_blank())

```

  There must be some good information in the state and geography data, 
  but plots like this are not very meaningful probably because the
  variation in state population, and hence the number of contributions 
  per state, dominates. Perhaps it would be better to focus on 
  means and/medians.

```{r echo=FALSE, warning=FALSE, Multivariate_Plots4}
# create a summary by state with some stats
campaign_gr_st_mm <- campaign %>% 
    group_by(contbr_st, party) %>%
    summarize(Mean   = mean(contb_receipt_amt),
              Median = median(contb_receipt_amt))

# melt in order to facet the plot by variable
campaign_gr_st_molten <- melt(campaign_gr_st_mm,
                              id.vars = c('contbr_st', 'party'))

# plot with states ordered by contribution size
ggplot(campaign_gr_st_molten, 
       aes(x = reorder(contbr_st, -value), 
           shape = variable) ) +
  geom_point(aes(y = value, 
                 color=party) ) +
  scale_color_manual(values = party_colors) +
  facet_wrap(~party, ncol = 1) +
  labs(title = "Means and Median Contribution by State", 
       x = "State", 
       y = "Per Contribution Amount in $") +
  theme1 +
  theme_x_state  + 
  theme(legend.position = 'bottom',
        legend.title = element_blank(),
        panel.grid.major = element_line(color = "white"))
```

  This chart overwhelmingly shows the broad Democratic base across all
  states, because both mean and median are relatively flat and the median 
  is very close to the median. By contrast The Republican 
  contribution size mean is substantially higher than the medians, with 
  substantial variation across the states.
  
  
  While there is substantial variation in mean contribution size by state, 
  the median is the same in about 80% of all states.

### Occupations

We will take a look at occupations. And see if there is some relationship. 
There are 
`r length(unique(campaign$contbr_occupation))` 
different occupations listed in this sample so let's look at the top 
occupations by frequency. First a chart of counts, then a chart of 
amounts because we have seen these differences before.

```{r echo=FALSE, warning=FALSE, Multivariate_Plots5A}
# Find top 30 ocurring occupations (filter missing) and calc the party percentages
top_occupations <- (campaign %>% 
  group_by(contbr_occupation) %>%
  summarize(count = n(),
            occ_dem_count = sum(party=="Democratic"),
            occ_rep_count = sum(party=="Republican"),
            occ_sum_amt = sum(contb_receipt_amt),
            occ_dem_amt = sum(ifelse(party=="Democratic", 
                                     contb_receipt_amt, 
                                     0)),
            occ_rep_amt = sum(ifelse(party=="Republican", 
                                     contb_receipt_amt, 
                                     0)),
            occ_dem_perc = round(100.0 * occ_dem_count / count, 1),
            occ_rep_perc = 100 - occ_dem_perc) %>%
  arrange(desc(count)) %>%
  filter(!(contbr_occupation %in% c("INFORMATION REQUESTED PER BEST EFFORTS",
                                    "INFORMATION REQUESTED", "RN"))) %>%
  mutate(occupation_rank = row_number())
  ) [1:30,] 

# plt the relative percentages 
ggplot (top_occupations, 
        aes(x = reorder(contbr_occupation, occ_dem_perc))) + 
  geom_bar(stat='identity', 
           aes(y = occ_rep_perc - occ_dem_perc,
              fill = ifelse(occ_rep_perc > occ_dem_perc, 
                            "Republican", 
                            "Democratic") ) ) +
  scale_fill_manual(name = "Party", values = party_colors) +
  scale_y_continuous(breaks = c(-100.0,  -50.0,  0.0, 50.0,  100.0), 
                     labels = c( "100%",  "50%", "0", "50%", "100%") ) +
  coord_flip() +
  labs(title = "Election 2012: Percentage difference in Contribution Counts\nTop occupations only",
       y = "Percentage Difference - Democratic vs. Republican", 
       x = "") +
  theme1 +
  theme(legend.position='bottom') +
  theme(legend.title = element_blank())

```

  The above chart shows the differences in percentage of
  contributions by occupations. Positive percentages are Republican - 
  Negative percentage are Democratic. The occupation order is 
  Democratic favoring to Republican favoring.
  
Next, contribution amounts

```{r echo=FALSE, warning=FALSE, Multivariate_Plots5B}
ggplot (top_occupations, 
        aes(x = reorder(contbr_occupation, occ_dem_perc))) + 
  geom_bar(stat='identity', 
           aes(y = (occ_rep_amt - occ_dem_amt) / 1E6,
           fill = ifelse(occ_rep_perc > occ_dem_perc, 
                        "Republican", 
                        "Democratic") ) ) +
  scale_fill_manual(values=party_colors) +
  scale_y_continuous(breaks = c(-1.0,  -0.5,  0.0, .5,  1.0), 
                     labels = c( "1",  "0.5", "0", "0.5", "1") ) +
  coord_flip() +
  labs(title = "Election 2012: Difference in Contribution Amount \nTop occupations only",
       y = "Total Contribution Difference in $ millions", 
       x = "") +
  theme1 +
  theme(legend.position='bottom',
        legend.title = element_blank())
```

  The above chart shows the differences in $ amounts of contributions by occupations. 
  Positive are Republican - Negative  are Democratic. The occupation 
  order is the same as the previous chart, Democratic favoring to Republican 
  favoring by count.
  
  A few observations:
  
  - The largest contribution total difference are from Professors, 
  Attorneys, Retired, and Homemakers
  - Fewer occupations dominate the Republican contributes 
  while the Democratic occupations are more varied. 
  `r sum(top_occupations$occ_rep_amt < top_occupations$occ_dem_amt )` of the 
  `r nrow(top_occupations)` occupations have more dollars 
  contributed to the Democratic Party.
  - Comparing to the previous chart, the crossover from Democratic to 
  Republican is much higher, which reflects the much larger 
  contribution amount prevalent with the Republican party.

### Compare Occupations by State 

Do retired people contribute to different parties depending on their state
of residence? This is the kind of question we will explore next.

```{r fig.width=9, fig.height=7, echo=FALSE, warning=FALSE, Multivariate_Plots5C}
campaign_gr_st_c <- campaign %>% 
    group_by(contbr_st, party) %>%
    summarize(count = n())
campaign_gr_st_perc <- dcast (campaign_gr_st_c, 
                           contbr_st ~ party, 
                           value.var = 'count')
campaign_gr_st_perc$dem_ratio <- campaign_gr_st_perc$Democratic /
                                 campaign_gr_st_perc$Republican
  

# merge campaign contributions with top_occupations to limit the occupations and assign rank
campaign_occ <- merge(campaign, 
                      top_occupations[,c('contbr_occupation',
                                         'occupation_rank',
                                         'occ_dem_perc', 
                                         'occ_rep_perc')])

# merge with state information
campaign_occ_st <- merge(campaign_occ, 
                         campaign_gr_st_perc[,c('contbr_st',
                                                'dem_ratio')])

# plot ordering so as to get Democratic top left and Republican bottom right
ggplot (campaign_occ_st,
        aes(x = reorder(contbr_st, -dem_ratio),
            y = reorder(contbr_occupation, -occ_rep_perc), 
            color = party, 
            size = contb_receipt_amt) ) +
  geom_jitter(alpha=0.2, 
              position = position_jitter(height=0, width=0)) +
  scale_color_manual(values = party_colors) +
  scale_size_continuous(name = "Per contribution amount $") +
  labs(title = "Election 2012: Contribution Amounts\nRepublican vs. Democratic",
       x = "State ordered by party\n <-- Democratic  -  Republican -->", 
       y = "Occupation ordered by party\n <-- Republican  -  Democratic -->" ) +
  guides(colour = guide_legend(override.aes = list(alpha=1, size=5)),
         size   = guide_legend(override.aes = list(alpha=0.5, size=5))) +
  theme_x_state +
  theme1 +
  theme(panel.background = element_rect(fill = 'white'),
        legend.position = 'bottom' )
```

  This chart turned out to be not very meaningful. The intent was 
  to chart the change in party leaning by Occupation by State. 
  By ordering the axes this yields a predominantly blue 
  Democratic presence top-left and red Republican presence 
  bottom-right. The state size influences the vertical bands which 
  distorts the information. 
  
  A few useful things can be seen. Retired people are heavy 
  contributors and split equally between the parties in all states. 
  Homemakers are clearly more likely to contribute to the Republican 
  party regardless of state. Professors and Educators are overwhelmingly 
  Democratic across all states, although their contributions vary in size. 
```

----

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The plots in this sectioned strengthened the idea that the size and number 
of contributions are very different for the Republican vs. the Democratic party. 

The final presidential candidate totals  tower over the other candidates. 
For example, Barack Obama total receipts are 
`r round(sum(subset(campaign$contb_receipt_amt, campaign$cand_nm == "Obama, Barack")) / sum(subset(campaign$contb_receipt_amt, campaign$cand_nm != "Obama, Barack"  & campaign$cand_nm != "Romney, Mitt")), 1)` 
times that of the total of all Republican candidates excluding Mitt Romney.

### Were there any interesting or surprising interactions between features?

Because swing states are pivotal in the election, 
I expected to find evidence of this. 
I quickly realized, however, that this can only 
be analyzed in the context of the state 
itself. States vary greatly in population and in 
per capital income. Perhaps these factors 
could be used in a future study to normalize the 
comparisons between states.

----

# Final Plots and Summary

### Plot ONE - How did fundraising progress for the Primary candidates?

This chart explores the growth -- cumulative contributions -- 
by month and year for the Republican participants in the 
presidential primary and omits the general 
election contenders, Mitt Romney and Barack Obama. 
The line width indicates the number (rate) of contributions. 
The original chart was hard to follow with colors - 
it took quite some time to figure out how to plot the names 
close to the lines.

```{r fig.width=9, echo=FALSE, Plot_One}
# summary by candidate to help with selection
# skip the general electon candidates (Romney & Obama)
# only dates up to the Republican convention (August 2012)
primary_gr_cand <- primary_gr_nm_ym %>%
              filter (isPrimaryCandidate(cand_nm) &
                      isPrimaryYm(contb_receipt_ym) ) %>%
              group_by(cand_nm) %>% 
              summarise(total_amt = max(cumsum_amt)) %>%
              arrange(desc(total_amt)) 
              

# merge the candidate names and totals with the summary by name (nm) and year-month (ym)
# include only "interesting" primary candidates
primary <- merge(primary_gr_cand,
                primary_gr_nm_ym,
                by = 'cand_nm')
primary$Contender <- ifelse(primary$total_amt >=200000, 
                            "Contender: $200,000 or more", 
                            "Non-contender: under $200,000")
# factor the name using descending total amount so the legend reads
# largest to smallest
primary$cand_nm  <- factor(reorder(primary$cand_nm, 
                                 - primary$total_amt))

# plot cumulative 
ggplot(primary) +
  geom_line(aes(x = contb_receipt_ym, 
                y = cumsum_amt/1E6, 
                group = cand_nm, 
                size = count,
                color = cand_nm ) ) +
  facet_grid ( ~ Contender) +
  scale_x_discrete(breaks = function (x) {breaks_yyyy_mm(x, every=6, ending_at=8)}, 
                  labels = labels_mmm_yy) +
  scale_y_continuous(breaks = pretty_breaks()) +
  scale_size_continuous(name = "Monthly Contribution Count\nis line thickness") +
  scale_color_manual(values = paired_palette,  name = "Candidate") + 
  labs(title = 'Election 2012: Cumulative Contributions\nPrimary Election Contenders vs. Non-contenders', 
       x = 'Month', 
       y = 'Contribution in $ millions') +  
  # soften the black size in the legend and make the colors thicker
  guides(colour = guide_legend(override.aes = list(size=2)),
         size =   guide_legend(override.aes = list(alpha=0.5))) +
  theme1 +
  theme (panel.margin = unit(20, "point"),
         strip.text = element_text(size = 12),
         legend.position = 'right')
```

  This chart focuses on the Republican primary candidates 
  who were real Contenders, raising $200,000 or more versus the 
  Non-contenders that never made much headway. Although the $200,000
  limit is somewhat arbitrary it does separate these groups into distinct
  shapes.
  
  The very interesting pattern on the left suggests candidates
  need a bump or surge in contributions to gain traction and stay in
  the race.  Note that Ron Paul, the most successful, is a little different, 
  starting early and having a more sustained climb rather than a sudden
  burst.
  
  Rick Perry's rapid rise in contribution amounts 
  from fewer contributors can be seen from the relatively thin line. 
  This suggests he may have been fueled by wealthy 
  contributors but unable to continue that into a sustainable 
  contribution base as can be 
  seen with the other long-lived candidates
  
  Ginrich and Santorum started later but grew steadily, 
  leveling out a little earlier than Ron Paul. 
  
  Ron Paul achieed his leadership wth far smaller contributions as can be seen 
  from the width of the line that indicates a larger number of contributions.
  
  It's also interesting to note that the slope of the rapid ascent period
  is very similar for most of the candidates, until the point 
  that they somewhat suddenly flatten.
  
----

### Plot TWO - Comparison of contribution size

This analysis has shown that there clearly are differences in the
size and number of contributions by party even though both parties
collected nearly similar total amounts. These charts place this stark difference side-by-side for easy comparison.

```{r fig.width=9, echo=FALSE, warning=FALSE, message=FALSE, Plot_Two}
# make a table using summarize for the  sums and mutate for the percentages
campaign_grouped_party <- campaign %>% 
  group_by(party, contb_receipt_amt_category) %>% 
  summarise(sum_amt = sum(contb_receipt_amt),
            count = n()) %>% 
  mutate ('Percent of Party Total Amount' = round(100 * sum_amt / 
                                                  sum(sum_amt),1),
          'Percent of Party Contribution Count' = round(100 * count   / 
                                                        sum(count)  ,1)) 
campaign_grouped_party_molten <- 
  melt(campaign_grouped_party[,c("party", 
                                 "contb_receipt_amt_category", 
                                 "Percent of Party Total Amount", 
                                 "Percent of Party Contribution Count")])
# plot the histogram
ggplot(campaign_grouped_party_molten,
       aes(x = contb_receipt_amt_category)) +
  geom_histogram(aes(y = value, 
                     fill = party),
                 stat = 'identity',
                 width = 0.6,
                 position = position_dodge()) + 
  facet_wrap (~variable) +
  coord_flip() +
  scale_y_continuous(breaks=seq(0,50,10)) +
  scale_fill_manual(values = party_colors) +
  labs(x ="Contribution Category in $", 
       y = "Percent of Party Total Contributions",
       title = "Election 2012: Contribution Amounts vs. Counts\nBy Party and Category") +
  theme1 +
  theme(legend.position ="bottom",
        legend.title = element_blank(),
        panel.margin = unit(20, "point"),
        strip.text = element_text(size = 12))

```

  From the left side chart above the difference in contribution 
  size is very clear. The Republican Party received 
  `r unlist(subset(campaign_grouped_party, party=='Republican' & contb_receipt_amt_category=='(1000,2500]', select="Percent of Party Total Amount"))` 
  percent of all funds from contributions in the range $1,000 to $2,500. 
  By contrast the Democratic Party received only 
  `r unlist(subset(campaign_grouped_party, party=='Democratic' & contb_receipt_amt_category=='(1000,2500]', select="Percent of Party Total Amount"))` percent 
  of all contributions from that category. In addition the Democratic 
  Party received more than 50% of all funds from contributions of $250 and under.
  
  From the right side chart, which shows the counts,  the broad base of 
  the Democratic Party is clear. About 60%  (35% + 25%) of Democratic 
  Contributions are $50 or under.

----

### Plot THREE- Democratic vs. Republican Contribution Pattern

 The total contributions for both parties 
 were very similar with the Republicans raising
`r round(100*sum(subset(campaign$contb_receipt_amt, campaign$party == "Republican")) / sum(subset(campaign$contb_receipt_amt, campaign$party == "Democratic")),1)`  % of the Democratic total.

Earlier exploratory charts suggested that the 
number and size of contributions are quite 
different between the two parties. The chart 
below extends this by exploring the timing, 
size and count of contributions for the two parties. 
Each contribution is plotted as a point 
in the traditional colors of the parties - Blue for the 
Democrats and Red for the Republicans. 
Choosing a very low alpha allows the distribution of the 
contributions to show through. Setting 
the size to be the contribution amount ensures that the 
color density overall reflects the relative 
value of each contribution. The horizontal bands are the size 
categories of the contribution. Time flows left to right, 

```{r echo=FALSE, Plot_Three}
# Compare amounts conntributed over time by party (excluding the green party)
ggplot (campaign, 
        aes(x = contb_receipt_ym, 
            y = contb_receipt_amt_category, 
            color = party, 
            size = contb_receipt_amt_category)) +
  geom_jitter(alpha = 0.02) +
  scale_color_manual(values = party_colors) +
  scale_x_yyyy_mm +
  labs(x = 'Month', 
       y = 'Contribution Category in $',
       title = 'Election 2012: Contribution Patterns\nRepublican vs. Democratic') +
  # low alpha causes legends to fade out, so override 
  guides(size = FALSE, 
         color = guide_legend(override.aes = list(alpha = 1, size=6))) +
  theme1 +
  theme(legend.position="bottom", 
        legend.background = element_rect( fill = 'white'),
        legend.title = element_blank(),
        panel.background=element_blank())
```


  This chart is quite striking and describes the differences between 
  the Democratic and 
  Republican fund raising. There are a few clear patterns:
  
  - Republicans make more donations above $250 and clearly lead 
  the $1000 - $2500 category. 
  
  - Democratic  contributions start earlier and seem more dominant in 
  the earlier stages except perhaps for December 2011 and January 2012. 
  This earlier start is curious since there was no primary competition 
  for the Democratic nomination.
  
  - The number of contributions in the few months before 
  the election in November, is dominated by Democrats, as can be seen 
  by the intense blue in May, June and July 2012.
  
  - In October, one month before the election,  Republicans 
  seem to suddenly increase the number and size of their contributions as 
  can be seen from the intense red.

----

# Reflection

This analysis took longer than expected because I felt compelled 
too look at many angles. 
In doing so I learned many aspects of R and ggplot 
that I otherwise wouldn't have.

### Struggles

- **Dataset size -** The size of the complete data set, as warned by our instructors,  slowed down the analysis, but it progressed more quickly once a sampling approach was taken. After a few samples matched the overall distribution well, the analysis proceeded with a 5% sample. (See the data munging document for details).

- **stats and ggplot -** I struggled with the use of stats with ggplot, and some review of those are definitely needed. However, in the later stage I used the dplyer  group_by, summarize "pipeline". Creating a result set
with the dplyr functions makes it much easier to verify the results
in table form before before plotting. Sums and totals can be checked 
interactively.

- **Raw dataset -** As warned by the instructors, figuring out the differences in the raw data took time, but was still interesting. 

### Lessons Learned

- **Perpetual tweaking - **"All results are interpretive so I found a continuous need to explore further. The results seemed that they could be refined and code could be continuously improved. Similarly, continuously tweaking the plots to adjust for minor formatting is very time-consuming, but essential to achieving visual clarity and simplicity.

- **Dates in many formats -**
  The usage of dates as months proved to be somewhat messy and I am not very comfortable with the end result. Most of the manipulation of dates is by month, which was represented as a yyyy-mm string, for example, '2012-08'. Although this worked reasonably well, it is treated as a factor and none of the
  specific continuous date scales can be used. In order to compare dates and use date scales, I assigned the first day of the month and used lubridate ymd function to create dates of type POSIXct. However, the first day of the month, such as "2012-08-01", which is really "2012-08-01 00:00:00 UTC" results in "2012-07-31 19:00:00" when used by the date scale functions which operate in local timezone.
  
    This can be seen in
    ```{r echo=TRUE}
       library(lubridate)
       strftime(ymd(paste("2012-08","01", sep="-")), "%F") 
    ```
  
    This was resolved by setting the date to the fifteenth of the month,
    which may be better because then the values align at about the 
    middle of the   month on the continuous scales. 
    
    It seems that almost every language I know has multiple 
    date manipulation
    libraries each with drawbacks and none having really solved all the 
    problems with dates.

### Successes

- **Broadbased contributions -**There is clear evidence in this data of the broader-base of contributions to the Democratic Party. While I understood this broadly I didn't know that this holds across almost every state.

- **Primary candidates -**The shapes of the contributions in the primary period, as shown in Plot 1, is particularly revealing. This certainly suggests that success in the primary could be predicted from the rate and size of contributions.

- **Occupation -**The variation by occupation is particularly interesting because this 
invokes a whole set of different personal questions such as:
Why do most writers contribute small amounts to the Democratic Party?

### Data restrictions, limitations

- **Sampling -**  Using a sample has precluded analysis by employer or individual and has limited the extent of analysis by occupation. It's possible that some of the state-specific counts by occupation may be very small in the sampled set and hence invalid. 

- **Business  contributions -** These contributions, with higher limits, were excluded to simplify and focus the analysis. 

- **Spousal contributions -** The data set includes transactions that exceed $2,500 and corresponding "transfers from spouse". This analysis does not handle these properly. 

- **Returns -** Campaign returns are excluded completely. We did estimate that returns only made up about 3% of the data set so the error of including a contribution but omitting the return in this 5% sample is ~ 3% * 5% = 0.15%

### Further Analysis

Key areas that could be pursued are:

- **Swing States -** Further analysis for swing states may be interesting, 
  relating contributions to population and income.

- **Business contributions -** Company contributions and contributions 
  by employees  may be very significant in total.
  It would be interesting to find employers with heavy 
  pockets of contributions.
  
- **Contributions to multiple candidates -**Do people contribute to
  the candidate they believe in or do they contribute to multiple candidates?

- **Swing states -**Further analysis of swing states will require 
eliminating the effect of state population. I would be interested 
in finding occupational differences in swing states that may indicate 
sets of people that become more active, the closer the race.

### Broader Meaning

In the 2012 election year, the Citizen's United 
decision cleared the way for much money 
to be contributed by corporations to PAC's and 
Super PACS (political action committee's), 
so this analysis only covers a small 
portion of the money involved in presidential elections.

Overall this analysis has reinforced the idea that the 
Democratic Party is more broadly based relying on 
smaller contributions from a broader base than the Republican Party. 



-----